---
title: "Assignment 5- Hockey Ballot"
author: Betsy Church
categories: [Assignment, DataViz]
code-fold: true
code-tools: true
---

# This assignment is about using data visualization to help choose what NHL Players are deserving of specific awards based on multiple different factors.

## Load Library and Environments

```{r, warning=FALSE}

library(tidyverse)
library(dplyr)
Goalies <-read.csv("C:/Users/betsy/OneDrive/Desktop/A5-TabularData/Goalies.csv")
RookieGoalies <-read.csv("C:/Users/betsy/OneDrive/Desktop/A5-TabularData/RookieGoalies.csv")
RookieSkaterIndividual <-read.csv("C:/Users/betsy/OneDrive/Desktop/A5-TabularData/RookieSkaterIndividual.csv")
RookieSkaterOnIce <- read.csv("C:/Users/betsy/OneDrive/Desktop/A5-TabularData/RookieSkaterOnIce.csv")
SkaterIndividualstats <-read.csv("C:/Users/betsy/OneDrive/Desktop/A5-TabularData/SkaterIndividualstats.csv")
SkaterOnIcestats <- read.csv("C:/Users/betsy/OneDrive/Desktop/A5-TabularData/SkaterOnIcestats.csv")

```

## Hart Memorial Trophy

Awarded to the “player judged most valuable to his team.” This isn’t necessarily the best overall player, but rather the one who contributes most significantly to his team’s success.

```{r}
players_data <- data.frame(SkaterOnIcestats)
  Player = c("Player")
  CF. = c("CF.")
  
players_data2 <-SkaterOnIcestats %>%
  select (Player, CF.) 

filtered_data <- players_data2 %>%
  dplyr::filter(CF. >= 65)




Hart<-ggplot(data=filtered_data, aes(x=CF., y= reorder(Player, CF.))) +
  geom_bar(stat="identity") +
  xlab("CF Percent")+
  ylab("Player Name")

print(Hart)
 
  


```

Based on the data here, with higher CF% meaning a more positive contribution to the team while on the ice, I would vote the following:

1.Samuel Buldoc

2.Nathan Clurman

3.Justin Sourdif

4.Shayne Gostisbehere

5.Rasmus Asblund

## Vezina Trophy

Presented to the goaltender “adjudged to be the best at this position.” NHL general managers vote on this award.

```{r}
vezina_trophy <-data_frame(Goalies)
 Player = c("Player")
  SV = c("SV.")
  
vez_data <-Goalies %>%
  select (Player, SV.) 

filtered_vez <-vez_data %>%
  dplyr::filter(SV. >=0.90)

  
ggplot(data=filtered_vez, aes(x=SV., y= reorder(Player, SV.))) +
  geom_bar(stat="identity") +
  xlab("SV")+
  ylab("Player Name")
  


```

This is great... But potentially misleading... Let's dig a bit deeper because who catches things 100% of the time? We want to know who has played the most games. So we are going to filter it to an arbitrary number of 45. This leaves us with 5 players.. Perfect! From there, let's look back at SV% with just those 5...

```{r}
vez_data1 <- data.frame(Goalies) %>%
   filter(GP >=45) %>%
   select(Player, SV.)


ggplot(data=vez_data1, aes(x=SV., y= reorder(Player, SV.)))+
geom_col()

```

Based on the SV% of these five players I am casting my ballot as follows:

1.  Connor Hellebuyck
2.  Andrei Vasilevskiy
3.  Ilya Sorokin
4.  Sam Montembeault
5.  Juuse Saros

## James Norris Memorial Trophy

Awarded to the defenseman who demonstrates “the greatest all-around ability” at the position.

```{R}
JN_trophy <-data.frame(SkaterIndividualstats) %>%
  filter(Position == "D") %>%
  filter(Shots.Blocked >=125) %>%
  select(Player, Shots.Blocked)

ggplot(data=JN_trophy, aes(x=Shots.Blocked, y= reorder(Player, Shots.Blocked)))+
geom_col()


```

This shows how many shots the individual blocked. Let's take a look at another aspect of defense and see if there are any overlaps!

```{R}
JN_trophy1 <-data.frame(SkaterIndividualstats) %>%
  filter(Position == "D") %>%
  filter(Takeaways >= 40) %>%
  select(Player,Takeaways)

  ggplot(data=JN_trophy1, aes(x=Takeaways, y= reorder(Player, Takeaways)))+
geom_col()

```

Hmm... There is only 1 name that overlaps and that is Jake Sanderson. So I will put him at the top of my list... I then have to decide what is more important; blocks or takeaways.... I can't choose so I am going to sprinkle in both and vote the following:

1.  Jake Sanderson
2.  Ian Cole
3.  Jaccob Slavin
4.  Jacob Trouba
5.  Cale Makar

## Calder Memorial Trophy

Given to the player “adjudged to be the most proficient in his first year of competition.” This is essentially the rookie of the year award.

```{r}
CM_Trophy <-data.frame(RookieSkaterOnIce) %>%
  filter(CF. >=60) %>%
  select(Player,CF.)

ggplot(data=CM_Trophy, aes(x=CF., y= reorder(Player, CF.)))+
geom_col()


```

Once again, I'm using CF% as my metric. Based on that, I would cast my ballot as follows:

1.  Nathan Clurman
2.  Samuel Bolduc
3.  Angus Crookshank
4.  Ryan Suzuki
5.  Alex Laferriere

## Frank J Selke Memorial Trophy

Awarded to the forward who best excels in the defensive aspects of the game.

I read that there are 3 forward positions: Center, Left wing and Right wing. I am going to look at Center position first. I am choosing to base the defensive aspect on blocking shots.

```{r}

FJS_Trophy_Center <-data.frame(SkaterIndividualstats) %>%
  filter(Position == "C") %>%
  filter(Shots.Blocked >= 60) %>%
  select(Player, Shots.Blocked)

ggplot(data=FJS_Trophy_Center, aes(x=Shots.Blocked, y= reorder(Player, Shots.Blocked)))+
geom_col()

```

I am now going to look at the Left wing

```{r}

FJS_Trophy_Left <-data.frame(SkaterIndividualstats) %>%
  filter(Position == "L") %>%
  filter(Shots.Blocked >= 60) %>%
  select(Player, Shots.Blocked)

ggplot(data=FJS_Trophy_Left, aes(x=Shots.Blocked, y= reorder(Player, Shots.Blocked)))+
geom_col()

```

Now I am looking at Right wing

```{r}
FJS_Trophy_Right <-data.frame(SkaterIndividualstats) %>%
  filter(Position == "R") %>%
  filter(Shots.Blocked >= 60) %>%
  select(Player, Shots.Blocked)

ggplot(data=FJS_Trophy_Right, aes(x=Shots.Blocked, y= reorder(Player, Shots.Blocked)))+
geom_col()


```

After looking at all of the forward positions and seeing who blocked greater than 60 shots I would cast my ballot as follows:

1.  Brandon Tanev
2.  Alex Tuch
3.  Noel Acciari
4.  Vincent Trocheck
5.  Rickard Rakell

## Lady Byng Memorial Trophy

Presented to the player who exhibits “the best type of sportsmanship and gentlemanly conduct combined with a high standard of playing ability.”

For this one, I plan to find the players with the lowest number of penalties, no misconduct, and the most points

```{r}
LBM_trophy <-data.frame(SkaterIndividualstats) %>%
  filter(Total.Penalties <= "10") %>%
  filter (Misconduct ==0) %>%
  filter (Total.Points >= 45) %>%
  select(Player, Total.Points)

ggplot(data=LBM_trophy, aes(x=Total.Points, y= reorder(Player, Total.Points)))+
geom_col()
```

Based on this criteria, I cast my ballot as follows:

1.  Kyle Connor
2.  Sidney Crosby
3.  Brayden Point
4.  Anze Kopitar
5.  Shea Theodore
