---
title: "Assignment_6_Hockey_Ballots_Winner"
Author: "Betsy Church"
Date: "March 23, 2025"
format: html
categories: [Assignment, DataViz]
code-fold: true
code-tools: true
---

## In this assignment, we are combining everyone's data to figure out who would win each award.

We start out by uploading the data set and getting our library packages all ready.

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)

CombinedNHL <- read.csv("C:/Users/betsy/OneDrive/CombinedNHL.csv")

```

## The Hart Memorial Trophy

```{r, warning=FALSE}

Hart1 <- CombinedNHL %>%
  select(Award, Rank, Player) %>%
  filter(Award == "Hart")

Weighted <-Hart1 %>%
  mutate(
    WeightedScore = case_when(
      Rank == 1 ~ 10,
      Rank == 2 ~ 7,
      Rank == 3 ~ 5,
      Rank == 4 ~ 3,
      Rank == 5 ~ 1
    )
  )

ggplot(data=Weighted, aes(x= WeightedScore,  y= reorder(Player, WeightedScore)))+
geom_col()+
  xlab("Weighted Score")+
  ylab("Player")
  
       

```

Based on the classroom votes, the winner of the Hart Trophy is **Leon Draisaitl**

## The Vezina Trophy

```{r, warning=FALSE}
Vez1 <- CombinedNHL %>%
  select(Award, Rank, Player) %>%
  filter(Award == "Vezina")

Weighted1 <-Vez1 %>%
  mutate(
    WeightedScore = case_when(
      Rank == 1 ~ 10,
      Rank == 2 ~ 7,
      Rank == 3 ~ 5,
      Rank == 4 ~ 3,
      Rank == 5 ~ 1
    )
  )

ggplot(data=Weighted1, aes(x= WeightedScore,  y= reorder(Player, WeightedScore)))+
geom_col()+
  xlab("Weighted Score")+
  ylab("Player")


```

Based on classroom votes, **Connor Hallebuyck** is the winner!

## The Calder Memorial Trophy

```{r, warning=FALSE}
Cal <- CombinedNHL %>%
  select(Award, Rank, Player) %>%
  filter(Award == "Calder")

Weighted2 <-Cal %>%
  mutate(
    WeightedScore = case_when(
      Rank == 1 ~ 10,
      Rank == 2 ~ 7,
      Rank == 3 ~ 5,
      Rank == 4 ~ 3,
      Rank == 5 ~ 1
    )
  )

ggplot(data=Weighted2, aes(x= WeightedScore,  y= reorder(Player, WeightedScore)))+
geom_col()+
  xlab("Weighted Score")+
  ylab("Player")


```

Based on our classroom vote, **Kirill Marchenko**

## The Frank J.Selke Trophy

```{r, warning=FALSE}
Selke <- CombinedNHL %>%
  select(Award, Rank, Player) %>%
  filter(Award == "Selke")

Weighted3 <-Selke %>%
  mutate(
    WeightedScore = case_when(
      Rank == 1 ~ 10,
      Rank == 2 ~ 7,
      Rank == 3 ~ 5,
      Rank == 4 ~ 3,
      Rank == 5 ~ 1
    )
  )

ggplot(data=Weighted3, aes(x= WeightedScore,  y= reorder(Player, WeightedScore)))+
geom_col()+
  xlab("Weighted Score")+
  ylab("Player")
```

Based on the classroom votes, the winner is **Brandon Tanev**

## The Lady Byng Memorial Trophy

```{r, warning=FALSE}
Byng<- CombinedNHL %>%
  select(Award, Rank, Player) %>%
  filter(Award == "Byng")

Weighted4 <-Byng %>%
  mutate(
    WeightedScore = case_when(
      Rank == 1 ~ 10,
      Rank == 2 ~ 7,
      Rank == 3 ~ 5,
      Rank == 4 ~ 3,
      Rank == 5 ~ 1
    )
  )

ggplot(data=Weighted4, aes(x= WeightedScore,  y= reorder(Player, WeightedScore)))+
geom_col()+
  xlab("Weighted Score")+
  ylab("Player")

```

Based on classroom votes, the winner is **Brayden Point**

## The James Norris Memorial Trophy

```{r, warning=FALSE}

Norris<- CombinedNHL %>%
  select(Award, Rank, Player) %>%
  filter(Award == "Norris")

Weighted5 <-Norris %>%
  mutate(
    WeightedScore = case_when(
      Rank == 1 ~ 10,
      Rank == 2 ~ 7,
      Rank == 3 ~ 5,
      Rank == 4 ~ 3,
      Rank == 5 ~ 1
    )
  )

ggplot(data=Weighted5, aes(x= WeightedScore,  y= reorder(Player, WeightedScore)))+
geom_col()+
  xlab("Weighted Score")+
  ylab("Player")


```

Based on classroom votes, the winner is **Cale Makar**
